---
title: "Analysis of Employee Earnings from CSO Data"
author: "Muhammad Raiyan 25210283"
format:
  pdf:
    latex-engine: xelatex
    documentclass: article
    classoption: portrait
    geometry: "top=2cm, bottom=2cm, left=2cm, right=2cm"
    fontsize: 11pt
    mainfont: "Times New Roman"
    linkcolor: black
    colorlinks: true
    toc: true          # <-- TURN TOC ON
    number-sections: true
    toc-depth: 2       # <-- ONLY # and ## headings in TOC
    header-includes:
      - \usepackage{fancyhdr}
      - \pagestyle{fancy}
      - \fancyhf{}
      - \fancyfoot[C]{\thepage}
      - \renewcommand{\headrulewidth}{0pt}
      - \renewcommand{\footrulewidth}{0pt}
      - \usepackage{titlesec}
      - \titlespacing*{\subsection}{0pt}{12pt}{6pt}
      - \usepackage{titling}
      - \setlength{\droptitle}{5cm}
      - \usepackage{titling}
      - \pretitle{\begin{center}\bfseries\huge}
      - \posttitle{\end{center}\vskip 0.5em}
      - \preauthor{\begin{center}\itshape}
      - \postauthor{\end{center}}
      - \predate{\begin{center}}
      - \postdate{\end{center}}
      
      

execute:
  echo: true
  warning: false
  message: false
editor: visual
---

\newpage

------------------------------------------------------------------------

# Part 1: Analysis of CSO Earnings Dataset

## 1. Overview

This analysis explores the **CSO Labour Market dataset** containing information on employment, average hourly earnings, and weekly paid hours for various types of employees across different economic sectors in Ireland.

-   **Categorical variables**: `economic_sector_nace_rev_2`, `type_of_employee`, `quarter`\
-   **Numerical variables**: `average_hourly_earnings`, `average_weekly_paid_hours`, `employment`

**Dataset source:**\
Central Statistics Office (CSO) Ireland. "Average Earnings, Hours Worked, Employment and Labour Costs (EHQ03)". [CSO Link](https://data.cso.ie/table/EHQ03)

```{r}
#| label: setup
#| echo: TRUE
# Load required packages
library(tidyverse)  # for data manipulation and plotting
library(purrr)      # required for the project

# Load the dataset
earnings <- read_csv("earnings_cso.csv")

# View first few rows
head(earnings)

```

### Interpretation:

The raw CSO dataset is in a long format: each row corresponds to one statistic (for example, “Average Hourly Earnings” or “Employment”) in a given quarter, sector and employee type. To make analysis easier, we will reshape it into a wide format where each statistic becomes its own column.

```{r}
#| label: pivot_data
#| echo: TRUE

library(tidyverse)

# Pivot long dataset to wide format
earnings_wide <- earnings |>
  pivot_wider(
    id_cols = c(Quarter, `Economic Sector NACE Rev 2`, `Type of Employee`),
    names_from  = `Statistic Label`,
    values_from = VALUE
  ) |>
  # Rename columns to simple names
  rename(
    quarter                    = Quarter,
    economic_sector_nace_rev_2 = `Economic Sector NACE Rev 2`,
    type_of_employee           = `Type of Employee`,
    average_hourly_earnings    = `Average Hourly Earnings`,
    average_weekly_paid_hours  = `Average Weekly Paid Hours`,
    employment                 = Employment
  ) |>
  # Convert categoricals to factors
  mutate(
    quarter                    = as.factor(quarter),
    economic_sector_nace_rev_2 = as.factor(economic_sector_nace_rev_2),
    type_of_employee           = as.factor(type_of_employee)
  ) |>
  # Keep only columns needed for Part 1
  select(
    quarter,
    economic_sector_nace_rev_2,
    type_of_employee,
    average_hourly_earnings,
    average_weekly_paid_hours,
    employment
  )

# Quick check
nrow(earnings_wide)
head(earnings_wide)





```

### Interpretation:

-   Pivoting reshapes the dataset so that each row now represents a combination of quarter, sector and employee type with all statistics stored in separate columns.

-   Renaming columns makes the code shorter and easier to read.

-   Converting categorical variables to factors ensures they are handled correctly in summaries and plots.

```{r}
#| label: summary_stats
#| echo: TRUE

library(knitr)  # for kable()

# Summary statistics of numeric variables
earnings_wide |>
  select(
    average_hourly_earnings,
    average_weekly_paid_hours,
    employment
  ) |>
  summary()

# Mean hourly earnings by sector
mean_by_sector <- earnings_wide |>
  group_by(economic_sector_nace_rev_2) |>
  summarise(
    mean_hourly_earnings = mean(average_hourly_earnings, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_hourly_earnings))

# Nicely formatted table
mean_by_sector |>
  kable(caption = "Mean hourly earnings by economic sector")
```

### Interpretation:

-   Average hourly earnings are **higher in Financial and Insurance activities**, while sectors like Accommodation and Food Services have lower wages.

-   Weekly hours are fairly consistent, with minor variation across employee types.

-   Employment numbers vary widely across sectors, with the highest counts in general industries and lower counts in sectors like Mining.

```{r}
#| label: boxplot_earnings
#| echo: TRUE
#| fig.cap: "Hourly Earnings by Employee Type"
#| fig.width: 10
#| fig.height: 9

# Remove rows with missing hourly earnings
earnings_box <- earnings_wide |>
  filter(!is.na(average_hourly_earnings))

# Boxplot of hourly earnings by employee type
p_box <- ggplot(
  data = earnings_box,
  aes(x = type_of_employee, y = average_hourly_earnings)
) +
  geom_boxplot(fill = "skyblue", width = 0.6, outlier.alpha = 0.6) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Hourly Earnings by Employee Type",
    x = "Employee Type",
    y = "Average Hourly Earnings (€)"
  ) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1),
    axis.title.x     = element_text(margin = margin(t = 10)),
    plot.title       = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )

p_box

```

### Interpretation:

-   Managers and professionals earn the **highest hourly wages**.

-   Clerical, sales, service, and manual workers have lower wages with less variability.

-   Boxplot reveals the **spread and median** of earnings across employee types.

```{r}
#| label: scatter_hours_earnings
#| echo: TRUE
#| fig.cap: "Relationship between weekly paid hours and hourly earnings by employee type"

# Remove any rows with missing values in the variables used
earnings_scatter <- earnings_wide |>
  filter(
    !is.na(average_weekly_paid_hours),
    !is.na(average_hourly_earnings)
  )

# Check there are rows left
nrow(earnings_scatter)

# Scatter plot of weekly paid hours vs hourly earnings
ggplot(
  data = earnings_scatter,
  aes(
    x = average_weekly_paid_hours,
    y = average_hourly_earnings,
    color = type_of_employee
  )
) +
  geom_point(alpha = 0.7, size = 2) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Hourly Earnings vs Weekly Paid Hours",
    x = "Average Weekly Paid Hours",
    y = "Average Hourly Earnings (€)",
    color = "Employee Type"
  ) +
  theme(
    plot.title       = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )

```

### Interpretation:

-   Each point represents a combination of sector, quarter and employee type, showing its weekly paid hours and hourly earnings.

-   The colouring by employee type helps to see whether some groups tend to work longer hours or earn higher wages.

-   The plot suggests that employee types with higher earnings generally do not have extremely low weekly hours, while some lower-earning groups may cluster at lower hourly earnings.

```{r}
#| label: purrr_summary
#| echo: TRUE

library(purrr)
library(knitr)  # for kable()

# Numeric columns
num_cols <- c("average_hourly_earnings",
              "average_weekly_paid_hours",
              "employment")

# Calculate mean of each numeric variable by sector
summary_by_sector <- earnings_wide |>
  group_by(economic_sector_nace_rev_2) |>
  summarise(
    across(all_of(num_cols), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )

# Nicely formatted table of means by sector
summary_by_sector |>
  kable(caption = "Mean earnings, hours and employment by economic sector")

# Column-wise mean across sectors using purrr
overall_means <- summary_by_sector |>
  select(all_of(num_cols)) |>
  map_dbl(mean)

overall_means

```

### Interpretation

-   The table summarises, for each sector, the average hourly earnings, weekly paid hours and employment.

-   Using purrr::map_dbl(), we compute the overall mean of each numeric variable across all sectors.

-   This demonstrates functional programming with purrr and provides a concise overview of typical earnings, hours and employment levels across the dataset.

### Summary of Part 1

-   The CSO earnings dataset was reshaped into a wide format and cleaned for analysis.

-   Numerical summaries and tables highlighted differences in average hourly earnings across sectors.

-   Boxplots and scatter plots visualised how earnings vary by employee type and weekly hours.

-   A purrr-based summary provided compact overall statistics, satisfying the requirement to use purrr.

\clearpage

------------------------------------------------------------------------

# Part 2: R Package Exploration

## 1. Overview

This part explores the **lubridate** R package, which provides convenient tools for working with dates and times. The aim is to show how lubridate can simplify common tasks such as converting character strings to dates and extracting components like month and weekday.

We first introduce the package and its authors, then create a small example dataset containing date strings and numeric values. Using this dataset, we:

-   convert the strings to proper `Date` objects with `ymd()`,
-   extract the month for each date using `month()`,
-   extract the weekday using `wday()`,
-   and compute a simple summary of the values by month.

These examples demonstrate the core functionality of lubridate and how it can be integrated into everyday data analysis workflows.

```{r}
#| Label: Lubridate_setup
#| echo: TRUE
library(lubridate)
citation("lubridate")

```

In words: the *lubridate* package (developed by Garrett Grolemund and Hadley Wickham) simplifies working with dates and times in R. It is available on **CRAN** at <https://CRAN.R-project.org/package=lubridate> .

We will demonstrate three key functions:

1.  ymd() – convert strings to Date objects,

2.  month() – extract the month from a date,

3.  wday() – extract the weekday from a date.

```{r}
# Example dataset with dates and some values
dates_df <- data.frame(
  date_string = c("2025-01-01", "2025-02-15", "2025-03-20", "2025-04-10"),
  value = c(10, 15, 12, 20)
)
dates_df

```

### Interpretation:

We create a small dataset with date strings and associated numeric values. These date strings will be converted to proper R Date objects so that we can easily extract components such as month and weekday.

```{r}
#| label: lubridate_ymd
#| echo: TRUE

# 1. Convert character strings to Date objects using ymd()
dates_df <- dates_df |>
  transform(
    date_parsed = ymd(date_string)
  )

dates_df


```

### Interpretation:

-   ymd() takes character strings in “year-month-day” format and converts them into R Date objects.

-   The new date_parsed column is now stored in a date format that R understands, which is essential for any time-based analysis.

```{r}
#| label: lubridate_components
#| echo: TRUE

# 2. Extract month (label = TRUE gives abbreviated names)
dates_df <- dates_df |>
  transform(
    month   = month(date_parsed, label = TRUE),
    weekday = wday(date_parsed, label = TRUE)
  )

dates_df
```

### Interpretation

-   **month()** extracts the month from a **Date**. With l**abel = TRUE**, we obtain abbreviated month names (e.g. “Jan”, “Feb”) instead of numbers.

-   **wday()** extracts the day of the week. With **label = TRUE**, we get short weekday names (e.g. “Mon”, “Tue”).

-   These functions make it easy to summarise data by month or weekday.

```{r}
#| label: lubridate_summary
#| echo: TRUE

# Average value by month
monthly_summary <- dates_df |>
  dplyr::group_by(month) |>
  dplyr::summarise(
    mean_value = mean(value),
    .groups = "drop"
  )

monthly_summary
```

### 

### Interpretation

-   Here we use the extracted month column to compute the average of value for each month.

-   This illustrates how the outputs of ymd(), month() and wday() can be combined with standard R tools (such as dplyr::summarise) to perform time-based summaries.

### Summary and Conclusion for Part 2

-   ymd() converts character strings into proper Date objects.

-   month() and wday() extract month and weekday information in a readable form.

-   Together, these lubridate functions make it much easier to manipulate and summarise time-based data in R.

\clearpage

------------------------------------------------------------------------

# Part 3: Custom Function and S3 Methods – Employee Analysis

## 1. Overview.

This section demonstrates how to create a custom R function to analyse the CSO earnings dataset.

-   We define a function employee_analysis() to summarise a numeric variable by employee type.

-   We assign an S3 class and implement custom print, summary and plot methods.

-   This satisfies the programming and S3 method requirements for Part 3.

```{r}
#| label: employee_function
#| echo: TRUE

# Custom function to summarise a numeric variable by employee type
employee_analysis <- function(data, numeric_var) {
  # Check that the chosen variable exists in the data
  if (!numeric_var %in% names(data)) {
    stop("Variable not found in dataset")
  }
  
  # Summary statistics by employee type
  summary_df <- data |>
    dplyr::group_by(type_of_employee) |>
    dplyr::summarise(
      mean   = mean(.data[[numeric_var]],   na.rm = TRUE),
      median = median(.data[[numeric_var]], na.rm = TRUE),
      sd     = sd(.data[[numeric_var]],     na.rm = TRUE),
      min    = min(.data[[numeric_var]],    na.rm = TRUE),
      max    = max(.data[[numeric_var]],    na.rm = TRUE),
      .groups = "drop"
    )
  
  # Create S3 object
  result <- list(
    variable = numeric_var,
    summary  = summary_df,
    data     = data
  )
  
  class(result) <- "employee_analysis"
  return(result)
}
```

### Interpretation

The function employee_analysis() takes a dataset and the name of a numeric variable of interest. It checks that the variable exists, computes mean, median, standard deviation, minimum and maximum by type_of_employee, and returns these results in an object of class "employee_analysis".

```{r}
#| label: employee_methods
#| echo: TRUE

# Print method for employee_analysis objects
print.employee_analysis <- function(x, ...) {
  cat("Employee Analysis - Variable:", x$variable, "\n")
  print(x$summary)
}

# Summary method: adds extra statistics (range, coefficient of variation)
summary.employee_analysis <- function(object, ...) {
  object$summary |>
    dplyr::mutate(
      range = max - min,
      coefficient_of_variation = sd / mean
    )
}

# Plot method: horizontal bar chart of mean values by employee type
plot.employee_analysis <- function(x, ...) {
  # Make a nicer label from the variable name (remove underscores)
  var_label <- gsub("_", " ", x$variable)

  ggplot(
    data = x$summary,
    aes(x = mean, y = type_of_employee, fill = type_of_employee)
  ) +
    geom_col(width = 0.6) +
    theme_minimal(base_size = 11) +
    labs(
      title = paste("Mean", var_label, "by Employee Type"),
      x = paste("Mean", var_label),
      y = "Employee Type"
    ) +
    theme(
      legend.position = "none",
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", hjust = 0.5)
    )
}

```

### Interpretation

-   print.employee_analysis() shows a short header followed by the basic summary table.

-   summary.employee_analysis() returns an enhanced table that adds the range (max – min) and the coefficient of variation (sd / mean) for each employee type.

-   plot.employee_analysis() produces a bar plot of mean values by employee type, formatted to be easy to read in the PDF.

```{r}
#| label: employee_example
#| echo: TRUE
#| fig.cap: "Mean average hourly earnings by employee type"

# Create an employee_analysis object
result <- employee_analysis(earnings_wide, "average_hourly_earnings")

# Show the basic summary in the console
print(result)

# Show the detailed summary table
summary(result)

# This line actually produces the plot in the document
plot(result)

```

### Interpretation

-   The employee_analysis() function is applied to average_hourly_earnings, producing an S3 object containing the summary by employee type.

-   The print() method displays the basic table of means, medians, standard deviations, minima and maxima.

-   The summary() method adds range and coefficient of variation, giving extra insight into variability within each employee group.

-   The plot() method visualises the mean hourly earnings across employee types in a clear bar chart.

### Summary of Part 3

-   A custom S3 class "employee_analysis" was created to summarise earnings by employee type.

-   Dedicated print, summary and plot methods provide different views of the results: textual summary, enhanced statistics, and a graphical display.

-   The example demonstrates how S3 methods can be used to build a small, coherent analysis tool in R.

------------------------------------------------------------------------

\newpage

# References

-   Central Statistics Office (CSO) Ireland – Earnings, Hours Worked, Employment and Labour Costs (EHQ03) <https://data.cso.ie/table/EHQ03>

-   Wickham, H. (2014). Tidy Data. Journal of Statistical Software, 59(10), 1–23. <https://doi.org/10.18637/jss.v059.i10>

-   Wickham, H., & Grolemund, G. (2017). R for Data Science. O’Reilly Media. <https://r4ds.had.co.nz>

-   Grolemund, G., & Wickham, H. (2011). lubridate: Make Dealing with Dates a Little Easier. CRAN R Package. <https://cran.r-project.org/package=lubridate>
